---
title: "HealthStats: Function Documentation"
author: "Nora Delgado"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

> **My Package:** `HealthStats`  
> **Dataset:** Personal wellness tracker (sleep, steps, heart‑rate, nutrition, stress, environment, mood)

---
  
## Why I’m Building *HealthStats*
I am passionate abaout personal wellnes and healthy lifestyle and I wonder if simple stats about daily habits could show how those habits link to someone's mood. Turning that idea into a mini‑R package felt like a good final‑project challenge.  `HealthStats` will:  
 - Clean the raw wellness sheet,  
 - Summarise the key health factors,  
 - Compare those factors by mood,  
 - Predict mood with a logistic model, and  
 - Show one daily "health score" so I can see good‑days vs bad‑days at a glance.

Everything below documents the **five core functions** my package will have. 
---
  
## Function #1 `clean_wellness_data()`
### Purpose
Make the raw sheet analysis‑ready by:
* dropping rows with essential NAs,
* standardising column names to snake case, and
* coercing character columns to factors.

### Implementation
```{r clean-fn}

clean_wellness_data <- function(input_path,
                                output_path = "wellness_data_cleaned.csv") {

# Load packages
  library(tidyverse)
  library(janitor)     
  
# Read data
  df <- read_csv(input_path, show_col_types = FALSE) %>% 
    clean_names()  # snake_case, strip weird chars
  
  # Split "systolic/diastolic" into two numeric cols
  df <- df %>% 
    separate(
      col   = systolic_and_diastolic_blood_pressure,
      into  = c("systolic_bp", "diastolic_bp"),
      sep   = "/",
      convert = TRUE,  
      remove  = TRUE
    )
  
  # Extract water (liters) & steps from text column
  # Regex finds "X liters" and "Y steps"
  df <- df %>% 
    mutate(
      water_liters = str_extract(
        additional_health_indicators_e_g_water_levels_steps,
        "[0-9.]+(?=\\s*liters?)"
      ) %>% as.numeric(),
      
      steps = str_extract(
        additional_health_indicators_e_g_water_levels_steps,
        "[0-9,]+(?=\\s*steps?)"
      ) %>% str_remove_all(",") %>% as.integer()
    ) %>% 
    select(-additional_health_indicators_e_g_water_levels_steps)  # drop original
  
  # Standardize text columns to lowercase / trimmed
  cat_cols <- c(
    "bmi_category",
    "quality_or_frequency_of_social_interactions",
    "activities_for_mental_health_such_as_therapy_sessions_or_meditation",
    "environmental_aspects_such_as_weather_and_air_quality",
    "lifestyle_decisions_such_as_drinking_and_smoking",
    "mood_output"
  )
  
  df <- df %>% 
    mutate(across(all_of(cat_cols), ~ str_trim(str_to_lower(.x))))
  
  # Ensure numeric columns really are numeric
  num_cols <- c(
    "duration_of_sleep_hours",
    "level_of_physical_activity_minutes_per_day",
    "level_of_stress_scale_1_10",
    "heart_rate_bpm",
    "level_of_workload_scale_1_10",
    "systolic_bp",
    "diastolic_bp",
    "water_liters",
    "steps"
  )
  
  df <- df %>% 
    mutate(across(all_of(num_cols), as.numeric))
  
  # Remove duplicates & rows missing critical vitals
  df <- df %>% 
    distinct() %>% 
    drop_na(duration_of_sleep_hours, heart_rate_bpm,
            systolic_bp, diastolic_bp)
  
  # Write cleaned dataset
  write_csv(df, output_path)
  
  message("Cleaned dataset written to: ", normalizePath(output_path))
  return(df)
}

```

### Example
```{r clean-example}
clean_data <- clean_wellness_data("C:/Users/norae/Downloads/wellness_data.csv")  
head(clean_data)
```

---
  
## Function #2 `summarize_health_factors()`
### Purpose
Quick numeric snapshot: average sleep, steps, heart‑rate, and water per day.

### Implementation
```{r summary-fn}
summarize_health_factors <- function(df) {
  df %>%
    summarize(
      avg_sleep       = mean(duration_of_sleep_hours),
      avg_steps       = mean(steps),
      avg_heart_rate  = mean(heart_rate_bpm),
      avg_water_intake= mean(water_liters)
    )
}
```

### Example
```{r summary-example}
summarize_health_factors(clean_data)
```

---
  
## Function #3 `summarize_by_mood()`
### Purpose
Show how habits shift across moods (Happy, Neutral, Sad).

### Implementation
```{r bymood-fn}
summarize_by_mood <- function(df) {
  df %>%
    group_by(mood_output) %>%
    summarize(
      avg_sleep = mean(duration_of_sleep_hours),
      avg_steps = mean(steps),
      avg_stress= mean(level_of_stress_scale_1_10),
      .groups = 'drop'
    )
}
```

### Example
```{r bymood-example}
summarize_by_mood(clean_data)
```

---
  
## Function #4 `predict_mood()`
### Purpose
A first‑cut logistic regression that guesses *Sad* vs *Not Sad* mood from sleep, steps, and stress.

### Implementation
```{r predict-fn}
predict_mood <- function(df) {
  df2 <- df %>%
    mutate(mood_binary = ifelse(mood_output == 'Sad', 1, 0))
  glm(mood_binary ~ duration_of_sleep_hours + steps + level_of_stress_scale_1_10,
      data   = df2,
      family = binomial)
}
```

### Example
```{r predict-example}
model <- predict_mood(clean_data)
summary(model)
```

---
  
## Function #5 `health_score()`
### Purpose
Roll sleep, steps, and water into one 0‑to‑1 daily score (weights: 40‑30‑30).

### Implementation
```{r score-fn}
health_score <- function(df) {
  df %>%
    mutate(
      health_score = 0.4 * (duration_of_sleep_hours / 8) +
        0.3 * (steps / 10000) +
        0.3 * (water_liters / 8)
    )
}

```

### Example
```{r score-example}
health_score(clean_data)
```